Supp mat: time series paper

### 2018-02-19 writing/manuscript_draft_short_version.tex
first update of short format version of the article

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{setspace}  
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}

@@

@@

\(^2\)Corresponding author:
\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}} , 1
480 203 7931 \vspace{3 mm}

Number of words (including 300 words per figure): 6,950 \vspace{1 mm}

Number of figures: 5 \vspace{1 mm}

Number of references: 34 \vspace{1 mm}

Supplementary material: 9 figures and 3 tables \vspace{3 mm}

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}).

To be submitted to: \emph{Biological Conservation} \vspace{3 mm}

Available as a preprint at: \url{https://peerj.com/preprints/3168/}
\vspace{3 mm}

Color does not need to be used for any figures in print \vspace{3 mm}

\vspace{3 mm}

\begin{spacing}{2}

@@

Observational studies and population time series have become a
cornerstone of modern ecological research and conservation biology
(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
to both understand population dynamics and to assess species extinction
risk. Even though many time series may now be considered ``long-term''
(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), most are
still short. Time series are typically short for a variety of reasons
(Field et al. 2007). They are often coupled with an experiment, which
may only last a couple of years. In addition, short funding cycles make
it difficult to examine populations over longer periods of time (Hughes
et al. 2017).

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). Scientists need to know the time series length required
to address a specific question. A short time series may lead to wrong
conclusions given large natural year-to-year variability (McCain,
Szewczyk, and Knight 2016). Managers need to know when action is needed
for a population. Therefore, managers must understand when population
trend over time is actually meaningful. For example, the International
Union for Conservation of Nature (IUCN) Red List Categories and Criteria
suggest, under Criterion A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or three generations (IUCN
2012). For both scientific and management questions, because sampling is
typically expensive, we also do not want to sample for longer than is
necessary. For example, Gerber, DeMaster, and Kareiva (1999)
investigated the minimum time series required to estimate population
growth of the endangered, but recovering, eastern North Pacific gray
whale (\emph{Eschrichtius robustus}). They used a long-term census to
retroactively determine the minimum time series required to assess
threat status. They found that only 11 years were needed, eight years
before the delisting decision was made. This highlights the importance
of estimating the minimum time series required as an earlier decision
would have saved time and money (Gerber, DeMaster, and Kareiva 1999).
Further, waiting too long to make a decision can imperil a species where
management action could have been taken earlier (Martin et al. 2012;
Martin et al. 2017).

An important step in experimental design is to determine the number of
samples required. For any particular experiment four quantities are
intricately linked: significance level (\(\alpha\)), statistical power,
effect size, and sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. A type I error is a false positive, or incorrect
rejection of a true null hypothesis. For example, if a time series was
assessed as significantly increasing or decreasing---when there was no
true significant trend---this would be a false positive. The false
positive rate, or significance level (\(\alpha\)) is often set at 0.05
(although this is purely historical, Mapstone (1995)). A type II error
(\(\beta\)) is a failure to detect a true trend, or failure to reject a
false null hypothesis. Formally, statistical power (\(1-\beta\)) is one
minus the probability of a type II error (\(\beta\)). The effect size is
a measure of the difference between two groups. Prior to an experiment,
one could set appropriate levels of power, significance level, and the
effect size to estimate the sample size required for the experiment.
This approach, however, is not straight-forward for a time series, or
more complicated scenarios (P. C. Johnson et al. 2015), as data are
clearly non-independent.

For time series data, two general approaches to estimating sample size
are appropriate. Simulations can be designed for a specific population
and question (Bolker 2008; P. C. Johnson et al. 2015). Simple models can
be simulated with parameter values corresponding to a population of
interest (Gerrodette 1987). Statistical power is the proportion of
simulations that meet some set of criteria. The specific criteria depend
on the question at hand. For example, given a time series, when is the
slope from linear regression significantly different from zero? In other
words, when is the time series significantly increasing or decreasing?
It is then possible to determine how power changes with a variable of
interest. For example, time series can be simulated for different
lengths of time. From these simulations, the minimum time series length
required to meet certain levels of statistical significance and power is
estimated (Bolker 2008).

In addition to using simulations, empirical time series can also be
used. Multiple replicates of similar populations are usually not
available, but it is possible to subsample an empirical time series
(Gerber, DeMaster, and Kareiva 1999; Brashares and Sam 2005). Subsamples
of different lengths can then be evaluated to estimate the proportion of
subsamples meeting some criteria, again a measure of statistical power.
Similar to the simulation approach, this measures of power can be used
to determine the minimum time series required for a particular question
of interest.

@@

I use both simulations and empirical time series to determine the
minimum number of years required to address several questions. I
estimate the minimum time series length required (\(T_{min}\)) to assess
long-term changes in abundance via simple linear regression. First, I
estimate \(T_{min}\) using a simulation approach. Then I examine 822
population time series to estimate \(T_{min}\). In the supplementary
material, I determine \(T_{min}\) for related ideas: using more
complicated population models, varying statistical level and power, and
the use of generalized additive models.

@@

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). I only briefly discuss this
approach as it has been described elsewhere. Essentially, a population
model is simulated repetitively for a number of years. This approach
requires us to determine values for model parameters (e.g.~population
variability). As an example, we can take the following population model
for population size \(N\) at time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
also the trend strength of the increase or decrease (i.e.~the rate of
increase). It is important to note that any population model could be
substituted for equation \ref{sim_model}, as in the supplementary
material (Figs. \ref{fig:min_time_dist_log_pop},
\ref{fig:shark_example}).

Statistical power is then the proportion of simulations that meet some
criteria. Here, our criteria is whether the slope parameter from linear
regression is significant at the 0.05 threshold with statistical power
of 0.8. Statistical power of 0.8 would indicate that, if there was a
true trend in abundance, there would be a 0.8 probability of detecting
the trend. Values of 0.05 for the significance level and 0.8 statistical
power are historical and it is important to examine the effect of
changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, a number of simulated time
series are shown for a set number of time periods (\(t=40\)). It is
clear that statistical power increases quickly with increases in length
of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is
greater than 0.8 (the dotted line), that is the minimum time required
(\(T_{min}\)) to be confident in the detection of a long-term trend in
abundance.

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) Example of a simulated time series for 40 time periods. (b)
Statistical power versus the simulated time series length. The
horizontal, dashed line is the desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve the
desired statistical power. (c) Minimum time required (\(T_{min}\)) for
simulations with different values of the trend strength (\(r\)). (d)
Minimum time required for different levels of population variability
(\(\sigma\)). In each case, the minimum time required is the minimum
number of years to achieve 0.8 statistical power given a significance
level of 0.05.\label{fig:theoretical_approach}}
\end{figure}

@@

I use a database of 2444 population time series compiled in (Keith et
al. 2015); they compared the predictability of growth rates among
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010) and several other
sources (Keith et al. 2015). I filtered out short time series (less than
35 years), and those with missing data, leaving 822 time series. The
data includes information on 477 vertebrate species with a focus on
mammals, birds, and fish. The data also includes information on
generation length and survey specifications. For each time series, I
also calculate other variables of interest: coefficient of variance in
population size, long-term trend in abundance (slope coefficient from
simple linear regression), and temporal autocorrelation. All analyses
were conducted in R (R Core Team 2016).

For a subset of populations (\(n =\) 547), there is information on
life-history traits available from another paper (Myhrvold et al. 2015),
including body size and generation length. All 547 populations were
birds. I examine how the minimum time required is related to these
life-history traits (Fig. \ref{fig:biological_correlates}).

@@

I assume that each time series is long enough to include all necessary
information (e.g.~variability) about the population. In other words,
each time series is a representative sample. I first take all possible
contiguous subsamples of each time series. For example, a time series of
35 years would have 34 possible contiguous subsamples of length 2, 34
possible contiguous subsamples of length 3, and continuing until 1
possible contiguous subsample of length 35 (Gerber, DeMaster, and
Kareiva 1999; Giron-Nava et al. 2017). Next, I run a linear regression
for each subsampled time series. Then, I determine the proportion of
subsamples of a particular length that have estimated slope coefficients
which are statistically different from zero. I only look at the
proportion of samples where the long-term, or ``true'', time series also
has a significant slope. This proportion is a measure of statistical
power. Lastly, I determine which subsample length is required to achieve
a certain threshold of statistical power (0.8, Cohen (1992)). The
minimum subsampled length that met these criteria is the minimum time
series length required (\(T_{min}\)).

In the supplementary material, I show how the same approach described
here for more complicated population models. I also determine the
minimum time required to estimate long-term trends according to
generalized additive models, instead of the simple linear models used
here (Fig. \ref{fig:gam_example}).

@@

I determined the minimum time series length (\(T_{min}\)) required to
address a particular question of interest. What is the minimum time
series length required to determine, via linear regression, the
long-term population trend? Here, the minimum time series length
required had high enough statistical power (greater than 0.8) for a set
significance level (\(\alpha\)) of 0.05. It is also possible to alter
statistical power and \(\alpha\). Predictably, with increased
statistical power or decreased \(\alpha\), \(T_{min}\) increased (Fig.
\ref{fig:min_time_vs_alpha_beta}). I then estimated \(T_{min}\) using
two approaches. I briefly describe results from the simulation approach
and then discuss the empirical approach.

\subsection{Simulation approach}\label{simulation-approach-1}

I constructed a general population model where the trend strength
(i.e.~slope coefficient) over time could be a model parameter. I then
simulated time series of different lengths. From these simulations I
determined the minimum time series length required to achieve a certain
level of statistical power. In line with past work (Gerrodette 1987), I
found the \(T_{min}\) increases (i.e.~more time is required) with
decreases in trend strength and with increases in population variability
(Figs. \ref{fig:theoretical_approach}c,d).

I chose a simple model, but any other population model could be used
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
I explored how the simulation approach could be applied to more
biologically-realistic population models (Fig. \ref{fig:shark_example}).
Specifically, I determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White, Nagy, and Gruber
2014). I found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Similar to
the simulation approach described above, the minimum time required for
the lemon shark population was strongly dependent on model parameters.

\subsection{Empirical approach}\label{empirical-approach-1}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{(a) Distributions of the minimum time required for populations
from four different biological classes. (b) Distribution of minimum time
required for all populations regardless of biological class. The minimum
time required calculation corresponds to a significance level of 0.05
and statistical power of 0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) coefficient of variation in
interannual population size, and (c) temporal lag-1
autocorrelation.\label{fig:correlates}}
\end{figure}

I examined a database of 822 separate population time series
representing 477 species. This database consisted of vertebrate species
with a variety of life-history characteristics (Fig.
\ref{fig:biological_correlates}). I limited analyses to populations with
at least 35 years of continuous sampling. I then examined the minimum
time required to estimate long-term trends via linear regression.

Across all the populations, I found an average minimum time series
length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
Birds (class Aves) had a much wider distribution of \(T_{min}\), but
usually required less years of sampling. Differences between these
classes can be explained by differences in variability in population
size and strength of trends in abundance (Fig. \ref{fig:class}).

\subsubsection{Corrrelates for minimum time
required}\label{corrrelates-for-minimum-time-required}

The minimum time series length required was strongly correlated with
trend strength (i.e.~estimated slope coefficient from linear
regression), coefficient of variation in population size, and
autocorrelation in population size (Fig. \ref{fig:correlates}). This is
in line based on simulations here and those of others (Rhodes and Jonzen
2011). Using a generalized linear model, with a Poisson error structure,
all three of these explanatory variables were significant and had large
effect sizes (see Table A1). Combined, trend strength, coefficient of
variation in population, and autocorrelation account for 75.1\% of the
explained deviance (Zuur et al. 2009) in minimum time series length
required.

For a subset of the populations I combined time series data with a data
on life-history characteristics of amniotes (Myhrvold et al. 2015).
There was life-history information available for 547 populations
representing 315 different species, all of which were birds (Aves
class).

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) log adult body mass (grams), (d) maximum longevity
(years), and (e) incubation (days). The lines in each plot represent the
best fit line from linear regression.\label{fig:biological_correlates}}
\end{figure}

Some life-history traits were significant predictors for the minimum
time required (Fig. \ref{fig:biological_correlates}, Tables A2,A3).
However, none of these life-history traits explained a large part of the
variation in minimum time required. In a generalized linear model, all
of the life-history traits described in figure
\ref{fig:biological_correlates} account for only 5.99\% of the explained
deviance in minimum time series length required. In addition, when
accounting for trend strength, coefficient of variation, and
autocorrelation, no life-history traits were significant predictors of
the minimum time required (Table A3).

\subsubsection{Evalulating the IUCN
criteria}\label{evalulating-the-iucn-criteria}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{Minimum time required to achieve 0.8 statistical power versus
the minimum time required under IUCN criteria A2 to classify a species
as vulnerable. Each point represents a single population, all of which
saw declines of 30\% or greater over a 10 year
period.\label{fig:IUCN_analysis}}
\end{figure}

The IUCN Red List Categories and Criteria suggest, under Criterion A2, a
species qualifies as vulnerable if it has experienced a 30\% decline
over 10 years, or 3 generations (whichever is longer) (IUCN 2012). I
examined a subset of populations with observed declines of 30\% or
greater over 10 years, qualifying all of them as vulnerable. This
resulted in n = 162 populations of fish, birds, and a single mammal. I
then compared the minimum time required to achieve 0.8 statistical power
(\(T_{min}\)) to the minimum time required under the IUCN criteria (Fig.
\ref{fig:IUCN_analysis}). For populations below the identity line in
figure \ref{fig:IUCN_analysis}, IUCN criteria would require more
sampling compared to estimates for \(T_{min}\). Further, populations
above the identity line are cases where the IUCN criteria would classify
a population as vulnerable despite not having sampled enough years to
achieve high statistical power (Fig. \ref{fig:IUCN_analysis}). The
silhouettes on figure \ref{fig:IUCN_analysis} highlight that species
with longer generation times typically have larger discrepancies between
\(T_{min}\) and the minimum time required for IUCN assessments (Fig.
\ref{fig:IUCN_correlates}).

\subsubsection{Sensitivity analysis}\label{sensitivity-analysis}

Lastly, I tested model sensitivity by using generalized additive models
(GAMs) instead of simple linear regression. Again, I examined the
minimum time required to estimate long-term population trends (Fig.
\ref{fig:gam_example}). I found that although I obtain a similar
distribution of minimum times required for GAMs, the minimum time
required for GAMs is on average shorter than for linear regression (Fig.
\ref{fig:min_time_dist_gam}).

@@

I explored two approaches to estimate the minimum time series length
required to address a particular question of interest. I asked, what is
the minimum time series length required to determine long-term
population trends using linear regression? This is one of the simplest
questions one could ask of a time series. The simulation-based approach
has been suggested by others, especially in situations more complicated
than that suited for classic power analysis (Gerrodette 1987; P. C.
Johnson et al. 2015; Bolker 2008). My simulations support past work that
longer time series are needed when the trend strength (i.e.~rate of
increase or decrease) is weak or when population variability is high
(Gerrodette 1987). I also showed how the simulation model can be altered
for a particular population (Fig. \ref{fig:shark_example}) or question
(Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).

Here, I focus on an empirical approach to estimate the minimum time
series length required to assess changes in abundance over time. I
examined 822 population time series (all longer than 35 years). I then
subsampled each to determine the minimum time required to achieve a
desired significance level and power for linear regression. Statistical
power is important as it provides on information as to the necessary
samples required to determine a significant trend (Legg and Nagy 2006).
I found that on average 15.91 years of continuous monitoring were
typically necessary (Fig. \ref{fig:min_time_dist}b). However, the
distribution of minimum time required was wide. This time-frame is in
line with past work on a short-lived snail species (Rueda-Cediel et al.
2015) and a long-lived whale species (Gerber, DeMaster, and Kareiva
1999). Hatch (2003) used seabird monitoring data to estimate minimum
sampling requirements. He found that the time required ranged from 11 to
69 years depending on species, trend strength, and study design.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}

\usepackage{titlesec}

\titlespacing\section{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
\titlespacing\subsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
\titlespacing\subsubsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}

@@

Running head: Power to detect population trends {[}33 of 40
characters{]}

@@

\(^2\)Corresponding author:
\href{mailto:eawhite@ucdavis.edu}{\nolinkurl{eawhite@ucdavis.edu}} , 1
480 203 7931 \vspace{1 mm}

To be submitted to: \emph{Ecology} as a Statistical Report \vspace{1 mm}

Number of pages: 20\\
\vspace{1 mm}

Number of figures: 4 \vspace{1 mm}

Number of references: 33 \vspace{1 mm}

Supplementary material: 9 figures and 3 tables \vspace{2 mm}

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}). \vspace{2 mm}

Available as a preprint at: \url{https://peerj.com/preprints/3168/}
\vspace{2 mm}

\pagebreak

@@

Observational studies and population time series have become a
cornerstone of modern ecological research and conservation biology
(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
to both understand population dynamics and to assess species extinction
risk. Some time series may now be considered ``long-term''
(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), but most
are still short. Time series are typically short due to short funding
cycles and typical experimental time-frames (Field et al. 2007; Hughes
et al. 2017).

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). A short time series may lead to wrong conclusions given
large natural year-to-year variability (McCain, Szewczyk, and Knight
2016). Managers need to know when action is needed for a population.
Therefore, managers must understand when a population trend over time is
actually meaningful. For example, the International Union for
Conservation of Nature (IUCN) Red List Categories and Criteria suggest,
under Criterion A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or three generations (IUCN
2012). Sampling is typically expensive, therefore, we also do not want
to sample for longer than is necessary. For example, Gerber, DeMaster,
and Kareiva (1999) investigated the minimum time series required to
estimate population growth of the endangered, but recovering, eastern
North Pacific gray whale (\emph{Eschrichtius robustus}). They used a
long-term census to retroactively determine the minimum time series
required to assess threat status. They found that only 11 years were
needed, eight years before the delisting decision was made. This
highlights the importance of estimating the minimum time series required
as an earlier decision would have saved time and money (Gerber,
DeMaster, and Kareiva 1999). Further, waiting too long to decide an
action can imperil a species where management action could have been
taken earlier (Martin et al. 2012; Martin et al. 2017).

@@

I estimate the minimum time series length required (\(T_{min}\)) to
assess long-term changes in abundance via simple linear regression.
Statistical power is not a new tool (Cohen 1992), but it is still
under-appreciated in ecological research (Legg and Nagy 2006).
Therefore, I only briefly describe the idea of statistical power for
time series analyses using simulations. I then examine 822 population
time series to estimate \(T_{min}\).

@@

\subsection{Statistical power in time
series}\label{statistical-power-in-time-series}

An important step in experimental design is to determine the number of
samples required. For any particular experiment four quantities are
intricately linked: significance level (\(\alpha\)), statistical power,
effect size, and sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. A type I error is a false positive, or incorrect
rejection of a true null hypothesis. For example, if a time series was
assessed as significantly increasing or decreasing---when there was no
true significant trend---this would be a false positive. The false
positive rate, or significance level (\(\alpha\)) is often set at 0.05
(although this is purely historical, Mapstone (1995)). A type II error
(\(\beta\)) is a failure to detect a true trend, or failure to reject a
false null hypothesis. Formally, statistical power (\(1-\beta\)) is one
minus the probability of a type II error (\(\beta\)). The effect size is
a estimate of the strength of a particular phenomenon. Prior to an
experiment, one could set appropriate levels of power, significance
level, and the effect size to estimate the sample size required for the
experiment. This approach, however, is not straight-forward for a time
series, or more complicated scenarios (P. C. Johnson et al. 2015), as
data are clearly non-independent.

@@

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). Essentially, a population
model, with a selected set of parameter values, is simulated
repetitively for a number of years. As an example, we can take the
following population model for population size \(N\) at time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth (\(r\)) is
the trend strength of the increase or decrease (i.e.~the estimated slope
from linear regression). It is important to note that any population
model could be used here. As an example, in the supplementary material
(Fig. \ref{fig:shark_example}), an age-structured model of population
dynamics is used in place of equation \ref{sim_model} (White, Nagy, and
Gruber 2014).

Statistical power is the proportion of simulations where the slope
parameter from linear regression is significant at the 0.05 threshold.
Statistical power of 0.8 would indicate that, if there was a true trend
in abundance, there would be a 0.8 probability of detecting the trend.
Values of 0.05 for the significance level and 0.8 statistical power are
historical (Cohen 1992) and it is important to examine the effect of
changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, the model in equation
\ref{sim_model} is simulated repetitively. Statistical power increases
with increases in the length of time sampled (Fig.
\ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the
dotted line), that is the minimum time required (\(T_{min}\)) to be
confident in the detection of a long-term trend in abundance. As shown
previously (Rhodes and Jonzen 2011; Rueda-Cediel et al. 2015),
statistical power increases with larger trend strength and lower
population variability (Fig. \ref{fig:theoretical_approach}c,d).
Simulation approaches can be useful before designing a monitoring
program or when a realistic model exists for the population in question.
However, it is also informative to examine how statistical power changes
with sampling effort in empirical time series.

@@

I use a database of 2444 population time series compiled by Keith et al.
(2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010) and several other
sources (Keith et al. 2015). I filtered out short time series (less than
35 years), and those with missing data, leaving 822 time series. The
data includes information on 477 vertebrate species with a focus on
mammals, birds, and fish.

@@

I assume that each time series is long enough to include all necessary
information (e.g.~variability) about the population. In other words,
each time series is a representative sample. I first take all possible
contiguous subsamples of each time series. For example, a time series of
35 years would have 34 possible contiguous subsamples of length 2, 34
possible contiguous subsamples of length 3, and continuing until 1
possible contiguous subsample of length 35 (Gerber, DeMaster, and
Kareiva 1999; Brashares and Sam 2005; Giron-Nava et al. 2017). In line
with the simulation approach, I determine the proportion of subsamples
of a particular length that have estimated slope coefficients
statistically different from zero. This proportion is a measure of
statistical power. Lastly, I determine which subsample length is
required to achieve a certain threshold of statistical power (0.8, Cohen
(1992)). The minimum subsampled length that met these criteria is the
minimum time series length required (\(T_{min}\)). All analyses were
conducted in R (R Core Team 2016).

@@

I examined a database of 822 separate population time series
representing 477 species. This database consisted of vertebrate species
with a variety of life-history characteristics. Across all the
populations, I found an average minimum time series length required
(\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution (Fig.
\ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
Birds (class Aves) had a much wider distribution of \(T_{min}\), but
usually required less years of sampling. Differences between these
classes can be explained by differences in variability in population
size and strength of trends in abundance (Fig. \ref{fig:class}).

\subsection{Corrrelates for minimum time
required}\label{corrrelates-for-minimum-time-required}

The minimum time series length required was strongly correlated with
trend strength (i.e.~estimated slope coefficient from linear
regression), coefficient of variation in population size, and
autocorrelation in population size (Figs. \ref{fig:correlates}a-c). All
three of these explanatory variables were significant and had large
effect sizes (see Table A1). Combined, trend strength, coefficient of
variation in population, and autocorrelation account for 75.1\% of the
explained deviance (Zuur et al. 2009) in minimum time series length
required.

There was life-history information available (Myhrvold et al. 2015) for
547 populations representing 315 different species, all of which were
birds (Aves class). Some life-history traits were significant predictors
for the minimum time required (Figs. \ref{fig:correlates}d-h, Tables
A2,A3). However, even combined, all five of the life-history traits
accounted for only 5.99\% of the explained deviance in minimum time
series length required. In addition, when accounting for trend strength,
coefficient of variation, and autocorrelation, no life-history traits
were significant predictors of the minimum time required (Table A3).

\subsection{Evalulating the IUCN
criteria}\label{evalulating-the-iucn-criteria}

I examined a subset of populations with observed declines of 30\% or
greater over 10 years, qualifying all of them as vulnerable under IUCN
Criterion A2 (IUCN 2012). This resulted in n = 162 populations. I then
compared the minimum time required to achieve 0.8 statistical power
(\(T_{min}\)) to the minimum time required under the IUCN criteria (Fig.
\ref{fig:IUCN_analysis}). For populations below the identity line in
figure \ref{fig:IUCN_analysis}, IUCN criteria would require more
sampling compared to estimates for \(T_{min}\). Further, populations
above the identity line are cases where the IUCN criteria would classify
a population as vulnerable despite not having sampled enough years to
achieve high statistical power (Fig. \ref{fig:IUCN_analysis}). The
silhouettes on figure \ref{fig:IUCN_analysis} highlight that species
with long generation times had larger discrepancies between \(T_{min}\)
and the minimum time required for IUCN assessments (Fig.
\ref{fig:IUCN_correlates}).

@@

I examined 822 population time series. I then subsampled each to
determine the minimum time required to achieve a desired significance
level and power for linear regression. Statistical power is important as
it provides information on the necessary number of samples required to
determine a significant trend (Legg and Nagy 2006). I found that on
average 15.91 years of continuous monitoring were typically necessary
(Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum
time required was wide. Therefore, simplistic rules of thumb are
probably not appropriate. This time-frame is in line with past work on a
short-lived snail species (Rueda-Cediel et al. 2015) and a long-lived
whale species (Gerber, DeMaster, and Kareiva 1999). Hatch (2003) used
seabird monitoring data to estimate minimum sampling requirements. He
found that the time required ranged from 11 to 69 years depending on
species, trend strength, and study design.

In line with theoretical predictions (Rhodes and Jonzen 2011), I also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to my prior expectations, I also found
that \(T_{min}\) did not correlate with any life-history traits (Fig.
\ref{fig:correlates}d-h). I initially hypothesized that species with
longer lifespans or generation times may require a longer sampling
period. This result could have been a result of at least two factors.
First, the data I used may not include a diverse enough set of species
with different life-history traits. Second, the question I posed,
whether a population is increasing or decreasing, was specifically
concerned with population trends over time. Therefore, life-history
characteristics may be more important for other questions more closely
tied to species biology. For example, Blanchard, Maxwell, and Jennings
(2007) used detailed simulations of spatially-distributed fisheries to
compare survey designs. They found that statistical power depended on
survey design, temperature preferences, and the degree of population
patchiness.
